Formalization of Forcing in Isabelle/ZF

### 2020-06-17 IJCAR2020/slides/forcing_IJCAR2020_slides.tex
almost final version of slides

\begin{frame}
  \frametitle{Why Isabelle/ZF?}
  \begin{block}{Pros}
    \begin{itemize}
    \item Most advanced set theory formalized (around 2017).
      % \item Closer to the mathematical practice of forcing.
    \item Structured proof language Isar
      \citep{DBLP:conf/tphol/Wenzel99}.
    \item Low in consistency strength.
    \end{itemize}
  \end{block}
  %
  \pause
  %
  \begin{block}{Cons}
    \begin{itemize}
    \item A fraction of automation of Isabelle (\texttt{sledgehammer},
      etc).
    \item Untyped, and too weak a metatheory.
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Isabelle/ZF \citep{DBLP:journals/jar/PaulsonG96}}
  \begin{itemize}
  \item<+-> An object logic of Isabelle axiomatized over the
    intuitionistic fragment  \emph{Pure} of higher order logic.
  \item<+-> It postulates \alert<3->{two types}: \isatt{i}
    (sets) and \isatt{o} (booleans). \visible<3->{\alert{Not
      inductively defined!}}
    \pause
  \item<+-> More stuff.
  \end{itemize}
\end{frame}

%% \begin{frame}
%%   \frametitle{\texttt{ZF-Constructible} \citep{paulson_2003}}
%%   Let $C :: {}$ \isatt{i} $\Rightarrow$ \isatt{o} be a predicate on sets.
%%   \pause
%%   \begin{block}{Relativization of $\sbq$}
%%     $x \sbq^C y \equiv \forall z.\ C(z) \longrightarrow (z\in x
%%     \longrightarrow z\in y)$.
%%   \end{block}
%%   \pause
%%   \begin{block}{$\sbq$ is absolute}
%%     $\forall x\ y.\ C(x) \land C(y) \longrightarrow (x \sbq^C y
%%     \longleftrightarrow x\sbq y)$.
%%   \end{block}
%%   \pause 
%%   \texttt{ZF-Constructible} develops relativized version of the axioms
%%   and proves that many concepts are absolute for \textbf{transitive
%%     classes} $C$
%%   \[
%%   x\in y \land C(y)  \longrightarrow C(x).
%%   \]
%%   satisfying additional axioms.
%% \end{frame}

@@

\begin{frame}
  \frametitle{The ctm approach to forcing, 1/4}
  \begin{block}{Countable transitive model (ctm) of $\ZF$}
    $\lb M, E\rb \models \ZF$ where
    \begin{itemize}
    \item \alert<3>{$M$ is \emph{standard}:\quad $E \defi {\in}\restriction M$.}
    \item $M$ is countable, and \emph{transitive}:\quad $x\in y \in M \implies x\in M$.
    \end{itemize}
  \end{block}
  %
  \pause%
  %
  \textbf{Note}. If $\lb N, R\rb \models \ZF$ with $R$ wellfounded, then there
  exists a ctm $M$ of $\ZF$.
  \pause

@@

  In this case, we know that  $\sbq$  is \textbf{absolute for
    transitive models}.
  %
  %% \begin{block}{Forcing notions and generics}
  %%   \begin{itemize}
  %%   \item $\lb \PP, {\preceq} ,\1\rb$ preorder with top.
  %%   \item A filter $G\sbq \PP$ is \textbf{$M$-generic} if it meets
  %%     every dense $D\sbq \PP$ in $M$.
  %%   \end{itemize}
  %% \end{block}
\end{frame}

@@

  Every $a\in M[G]$ is coded by
  some $\punto{a}\in M$ through the $M$-definable function $\val$:
  \[
  M[G]\defi \{ \val(G,\punto{a}) : \punto{a}\in M\}
  \]
  Fundamentally, \textbf{truth} in $M[G]$ is coded in $M$ by the
  function $\forceisa$.
  \pause
  \begin{theorem}[\citep{Cohen-CH-PNAS}]
    \begin{enumerate}
    \item $\bigl(\forall G\visible<4->{\alert{\ni p}}.\quad M[G], [\val(G,\punto{a})]
      \models \phi\bigr)
      \iff
      M,
      [\only<-3>{\1}\alert{\only<4->{p\hspace{0.2ex}}},\PP,\preceq,\punto{a}]\models
      \forceisa(\phi)$.
    \item $\forceisa$ is definable over $M$.\phantom{$\punto{a}).\models \phi\bigr)
      \iff M,$}\visible<4->{%
      $\overset{\searrow}{}$ \ 
      $\alert{p} \forces^M_{\PP,\preceq} \phi(\punto{a})$
      \  $\overset{\swarrow}{}$}
      \pause% 
      \pause% 
    \item For $env\in\mathsf{list}(M)$,
      \[
      M[G], \mathrm{map}(\val(G),env) \models \phi
      \iff
      \exists p\in G.\ M, [p,\preceq,\PP] \mathbin{{+}\!\!\!\!{+}} env \models
      \forceisa(\phi).
      \]
    \end{enumerate}
  \end{theorem} 
\end{frame}

@@

\begin{frame}{Some details}
  \begin{enumerate}\setcounter{enumi}{1}
  \item<+-> We formalized the formula transformer $\forceisa$ and
    hence the forcing relation $\forces$, and proved the Fundamental
    Theorems.
  \end{enumerate}
  \begin{itemize}[I]
  \item<+-> $\forceisa$ is defined by induction on internalized
    formulas.
  \item<+-> The base cases  $\forceisa(x\in y)$ and  $\forceisa(x =
    y)$ are  defined by wellfounded
    recursion, which \alert{must also be internalized}.
  \end{itemize}
\end{frame}

@@

\begin{frame}<1-12>[label=code-review]
  \begin{center}\setlength{\fboxsep}{0pt}
    \only<01>{\includegraphics[draft,scale=0.23]{kunen_powerset_01.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_01.png}}}%
    \only<02>{\includegraphics[draft,scale=0.23]{kunen_powerset_02.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_02.png}}}%
    \only<03>{\includegraphics[draft,scale=0.23]{kunen_powerset_03.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_03.png}}}%
    \only<04>{\includegraphics[draft,scale=0.23]{kunen_powerset_04.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_04.png}}}%
    \only<05>{\includegraphics[draft,scale=0.23]{kunen_powerset_05.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_05.png}}}%
    \only<06>{\includegraphics[draft,scale=0.23]{kunen_powerset_06.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_06.png}}}%
    \only<07>{\includegraphics[draft,scale=0.23]{kunen_powerset_07.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_07.png}}}%
    \vspace{3.00115pt}\only<08>{\includegraphics[draft,scale=0.23]{kunen_powerset_08.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_08.png}}}%
    \only<09>{\includegraphics[draft,scale=0.23]{kunen_powerset_09.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_09.png}}}%
    \only<10>{\includegraphics[draft,scale=0.23]{kunen_powerset_10.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_10.png}}}%
    \only<11>{\includegraphics[draft,scale=0.23]{kunen_powerset_11.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_11.png}}}%
    \only<12>{\includegraphics[draft,scale=0.23]{kunen_powerset_12.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_12.png}}}%
    \only<13>{\includegraphics[draft,scale=0.23]{kunen_powerset_13.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_13.png}}}%
    \only<14>{\includegraphics[draft,scale=0.23]{kunen_powerset_14.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_14.png}}}%
    \only<15>{\includegraphics[draft,scale=0.23]{kunen_powerset_15.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_15.png}}}%
    \only<16>{\includegraphics[draft,scale=0.23]{kunen_powerset_16.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_16.png}}}%
    \only<17>{\includegraphics[draft,scale=0.23]{kunen_powerset_17.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_17.png}}}%
    \only<18>{\includegraphics[draft,scale=0.23]{kunen_powerset_18.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_18.png}}}%
    \only<19>{\includegraphics[draft,scale=0.23]{kunen_powerset_19.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_19.png}}}%
    \only<20>{\includegraphics[draft,scale=0.23]{kunen_powerset_20.png} \\ \fbox{\includegraphics[draft,scale=0.26]{isa_converted_20.png}}}%
  \end{center}
\end{frame}

@@

\begin{frame}
  \frametitle{Looking forward}
  \begin{block}{Formalizing math}
    \begin{itemize}
    \item<+-> Cofinality, K\H{o}nig's Theorem, Shanin's $\Delta$-system Lemma.
    \item<+-> Forcing notion for adding $\kappa$ Cohen reals.
    \item<+-> Theorems on preservation of cardinals.
    \end{itemize}
  \end{block}
  %
  \begin{block}{Technical aids}<+->
    \begin{itemize}
    \item<.-> Automatic relativization and proof of absoluteness of concepts.
    \item<+-> ``Relative functions'' (e.g., $\P^M$, $\card{\cdot}^M$, $\cf^M$).
    \end{itemize}
  \end{block}
\end{frame}

>>>>

\begin{frame}
  \frametitle{Why Isabelle/ZF?}
  \begin{block}{Pros}
    \begin{itemize}
    \item Most advanced set theory formalized (around 2017).
      % \item Closer to the mathematical practice of forcing.
    \item Structured proof language Isar
      \citep{DBLP:conf/tphol/Wenzel99}.
    \item Low in consistency strength.
    \end{itemize}
  \end{block}
  %
  \pause
  %
  \begin{block}{Cons}
    \begin{itemize}
    \item A fraction of automation of Isabelle (\texttt{sledgehammer},
      etc).
    \item ``Untyped'', and too weak a metatheory.
    \end{itemize}
  \end{block}
\end{frame}

\begin{frame}
  \frametitle{Isabelle/ZF \citep{DBLP:journals/jar/PaulsonG96}}
  \begin{itemize}
  \item<.-> An object logic of Isabelle axiomatized over the
    intuitionistic fragment  \emph{Pure} of higher order logic.
  \item<+-> It postulates two types: \isatt{i}
    (sets) and \isatt{o} (booleans). \visible<3->{\alert{Not
      inductively defined!}}
  \item<+-> The Replacement and Separation axiom schemes feature free
    high order variables.
  \item<+-> Induction/recursion is \emph{internal} to the theory (it works as a
    layer on top of set-theoretical proofs of well-foundedness (of
    $\N$, of $\Ord$, etc).
  \end{itemize}
\end{frame}

@@

\begin{frame}
  \frametitle{The ctm approach to forcing, 1/4}
  \begin{block}{Countable transitive model (ctm) of $\ZF$}
    $\lb M, E\rb \models \ZF$ where
    \begin{itemize}
    \item \alert<3>{$M$ is \emph{standard}:\quad $E \defi {\in}\restriction M$.}
    \item $M$ is countable, and \emph{transitive}:\quad $x\in y \in M \implies x\in M$.
    \end{itemize}
  \end{block}
  %
  \pause%
  %
  \textbf{Note}. If $\lb N, R\rb \models \ZF$ with $R$ well-founded, then there
  exists a ctm $M$ of $\ZF$.
  \pause

@@

  In this case, we know that  $\sbq$  is \textbf{absolute for
    transitive models}.
\end{frame}

@@

  Every $a\in M[G]$ is coded by
  some $\punto{a}\in M$ through the function $\val$:
  \[
  M[G]\defi \{ \val(G,\punto{a}) : \punto{a}\in M\}
  \]
  Fundamentally, \textbf{truth} in $M[G]$ is coded in $M$ by the
  function $\forceisa$.
  \pause
  \begin{theorem}[\citep{Cohen-CH-PNAS}]
    \begin{enumerate}
    \item $\forceisa$ is definable over $M$.
      %
    \item For every $\phi$, $M$-generic $G$, and $\punto{a}\in M$,
      \[
      M[G], [\val(G,\punto{a})] \models \phi
      \quad      \iff  \quad 
      \exists p\in G.\ \; M, [p,\preceq,\PP, \punto{a}]  \models
      \forceisa(\phi).
      \]
      \pause
      \rule{21em}{0pt}
      \raisebox{1.6ex}[0ex]{$\overset{\searrow}{}$ \ \ 
        $p \alert{\forces^M_{\PP,\preceq}} \phi(\punto{a})$
        \  \ $\overset{\swarrow}{}$}
    \end{enumerate}
  \end{theorem} 
\end{frame}

@@

\begin{frame}{Some details}
  \begin{enumerate}\setcounter{enumi}{1}
  \item We formalized the formula transformer $\forceisa$ and
    hence the forcing relation $\forces$, and proved the Fundamental
    Theorems.
  \end{enumerate}
  %
  \pause
  %
  \begin{block}{$\forceisa$ is defined by recursion on  formulas}
    This is the main reason we work with the set of internalized
    formulas, and that we require legit first-order expressions for
    the axiom schemes (Separation and Replacement).

    The base cases  $\forceisa(x\in y)$ and  $\forceisa(x =
    y)$ are  defined by \alert{internal} well-founded
    recursion.
  \end{block}
  %
  \pause
  \begin{alertblock}{}
    We enhanced the recursion results of Isabelle/ZF as well as
    the relevant preservation results in ZF-Constructible, thus
    showing that forcing is absolute for atomic formulas.
  \end{alertblock}
\end{frame}

@@

\begin{frame}<1-12>[label=code-review]
  \begin{center}\setlength{\fboxsep}{0pt}
    \only<01>{\includegraphics[scale=0.23]{kunen_powerset_01.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_01.png}}}%
    \only<02>{\includegraphics[scale=0.23]{kunen_powerset_02.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_02.png}}}%
    \only<03>{\includegraphics[scale=0.23]{kunen_powerset_03.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_03.png}}}%
    \only<04>{\includegraphics[scale=0.23]{kunen_powerset_04.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_04.png}}}%
    \only<05>{\includegraphics[scale=0.23]{kunen_powerset_05.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_05.png}}}%
    \only<06>{\includegraphics[scale=0.23]{kunen_powerset_06.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_06.png}}}%
    \only<07>{\includegraphics[scale=0.23]{kunen_powerset_07.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_07.png}}}%
    \vspace{3.00115pt}\only<08>{\includegraphics[scale=0.23]{kunen_powerset_08.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_08.png}}}%
    \only<09>{\includegraphics[scale=0.23]{kunen_powerset_09.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_09.png}}}%
    \only<10>{\includegraphics[scale=0.23]{kunen_powerset_10.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_10.png}}}%
    \only<11>{\includegraphics[scale=0.23]{kunen_powerset_11.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_11.png}}}%
    \only<12>{\includegraphics[scale=0.23]{kunen_powerset_12.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_12.png}}}%
    \only<13>{\includegraphics[scale=0.23]{kunen_powerset_13.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_13.png}}}%
    \only<14>{\includegraphics[scale=0.23]{kunen_powerset_14.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_14.png}}}%
    \only<15>{\includegraphics[scale=0.23]{kunen_powerset_15.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_15.png}}}%
    \only<16>{\includegraphics[scale=0.23]{kunen_powerset_16.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_16.png}}}%
    \only<17>{\includegraphics[scale=0.23]{kunen_powerset_17.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_17.png}}}%
    \only<18>{\includegraphics[scale=0.23]{kunen_powerset_18.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_18.png}}}%
    \only<19>{\includegraphics[scale=0.23]{kunen_powerset_19.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_19.png}}}%
    \only<20>{\includegraphics[scale=0.23]{kunen_powerset_20.png} \\ \fbox{\includegraphics[scale=0.26]{isa_converted_20.png}}}%
  \end{center}
\end{frame}

@@

\begin{frame}
  \frametitle{Looking forward}
  \begin{block}{Formalizing math}
    \begin{itemize}
    \item Cofinality, K\H{o}nig's Theorem, Shanin's $\Delta$-system Lemma.
    \item Forcing notion for adding $\kappa$ Cohen reals.
    \item Theorems on preservation of cardinals.
    \end{itemize}
  \end{block}
  \pause
  \begin{block}{Technical aids}
    \begin{itemize}
    \item Automatic relativization and proof of absoluteness of concepts.
    \item ``Relative functions'' (e.g., $\P^M$, $\card{\cdot}^M$, $\cf^M$).
    \end{itemize}
  \end{block}
\end{frame}
